Cultural Evolution of Conservation Behaviors Within and Between Communities

### 2020-12-16 DissertationProposal/DissertationProposal.Rmd
done for the day. Forever?

--- 
title: Cultural Evolution of Conservation Behavior at the Intersection of Working
  Landscapes and Protected Areas
author: "Matt Clark"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: yes
  pdf_document:
    toc: yes
    extra_dependencies: "subfig"
  bookdown::html_document2:
    toc: yes
fontsize: 12pt
link-citations: yes
csl: AmJBot.csl
subtitle: Dissertation Proposal for Committee Review
bibliography: [References/Dissertationproposal.bib,References/NonZoteroRefs.bib]
---
```{r packages, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(rmarkdown)
library(bookdown)
library(formattable)
library(kableExtra)
library(dplyr)
library(magrittr)
library(prettydoc)
library(htmltools)
library(knitcitations)
library(devtools)
library(tidyverse)
library(scales)
library(extrafont)
```

@@

# Chapter 2: Effects of inter-group resource competition and selection on community adoption of costly conservation policy.

## Background
Human social structures developed and still exist in nested groups, ranging from a nuclear family unit to a national or even multi-national identity. Group behavioral norms can have profound impacts on the wellbeing of group members and their environment, far beyond that of a single individual. Cultural group selection theorizes, and to a small extent has empirically demonstrated, the mechanisms which establish and perpetuate group-level norms [@henrich_cultural_2004; @bell_culture_2009]. Most relevant to community conservation is the development of norms which are beneficial for the group, but cause an individual to incur some personal cost; we refer to these as costly cooperative behaviors [@molleman_cultural_2013]. Central to theories around the adoption of costly cooperative behaviors is the insight that selection acts on both individuals and groups. Individuals may engage in personally costly behavior if it is beneficial for their group, which they also depend on for their wellbeing [@handley_human_2020]. The degree to which an individual depends on each of their nested groups is directly proportional to the personal costs they will incur for that group [@okasha_multilevel_2004; @price1972extension]. To the extent that conservation is a collaborative dilemma, these insights from cultural group selection may be paramount to implementing effective community based conservation [@waring2015multilevel; @kline2018designing].

@@

REDD+ in Zanzibar was designed to give performance based payments to communities who successfully reduce deforestation [@burgess2010getting]. In theory, these immediate, promised payments should alter the ratio of individual costs to group benefits so that costly cooperation is favored under cultural group selection. While this project was implemented in only a select few communities (Figure \@ref(fig:moon)), this setup should also promote the spread of community conservation norms to adjacent communities through success-biased imitation [@boyd_transmission_2010].

The REDD+ program on Pemba helped establish CoFMAs in 18 wards on the island, some of which have been successful in slowing and even stopping deforestation (Figure \@ref(fig:ForestCoverMap). The payments for the conservation efforts however were never given to communities. While this is ethically troublesome, it is interesting in light of the fact that ten additional wards on the island have subsequently elected to establish CoFMAs. This leaves open the question of why community conservation areas are spreading on the island while they are not expected to do so under cultural group selection. One explanation posited by Andrews and Borgerhoff Mulder [-@andrews_cultural_2018] actually hinges on a major critique of REDD+; leakage [@andam2008measuring].

Leakage [as defined by @andam2008measuring] is a form of land use spillover where an intervention, such as the CoFMAs established on Pemba by REDD+, has a negative, counteracting effect on the targeted outcome variable in another place. On Pemba, we have anecdotal testimonies that individuals in wards with CoFMAs have simply shifted their harvest pressure onto adjacent wards. While dubious for the efficacy of REDD+ writ large, on Pemba, leakage may actually be driving the spread of community based conservation observed in the ten new communities which elected to create CoFMAs post REDD+ [@andrewsdoes]. Just as the performance based payments from REDD+ were designed to alter the individual/community cost/benefit matrix so that selection favors costly collaboration via immediate community payments, we hypothesize that leakage from wards with CoFMAs may also act to favor costly cooperative behavior as individuals must act to defend their access to resources they would not have otherwise.

@@

### Predictions
* P1: Regardless of the demarcated size of the conservation area in each CoFMA, the leakage pressure on adjacent wards will be associated with the rate of deforestation, i.e. the extent to which individuals adhere to conservation area regulations.

* P2: In wards that have subsequently applied for CoFMAS, the primary motivator will be to combat leakage from adjacent wards, more so than copying their success or intrinsic motivations.

## Research Design
Under cultural group selection theory, clear group boundaries must be present for selection to act at the group level. These group level selection pressures are what we hypothesize to drive the adoption of CoFMAs. Logically, groups with more clear differences from surrounding groups should be more likely to adopt CoFMAs than groups whose boundaries with surrounding groups are blurred. For this reason, I will calculate the cultural $F_{ST}$ values for all combinations of neighboring Pemban wards as opposed to pairwise calculations of cultural $F_{ST}$ across all wards overall.

I will conduct an in-person survey of randomly sampled forest users in each Pemban ward. This survey will elicit information on approximately 50 cultural norms ranging from crime, cultural markers, punishment, family dynamics, etc. Cultural $F_{ST}$ values will be calculated for each trait given the following formula:

$$F_{STx} = \frac{\frac{n_{i}}{n_{i}+n_{j}} (p_{i}-\bar{p})^2 + \frac{n_{j}}{n_{i}+n_{j}} (p_{j}-\bar{p})^2}{\bar{p}(1-\bar{p})}$$

Where $i$ and $j$ are the two populations being compared, $n_{i}$ and $n_{j}$ represent the number of survey respondents for each group respectively. For trait $x$, $\bar{p}$ is the total frequency of trait $x$ in both populations, $p_{i}$ and $p_{j}$ are the trait frequency in each population. The equation can be interpreted as the variance in the relative frequency of trait $x$ between populations divided by the total variance in the trait across both populations. These values will be averaged across all traits to produce the total cultural $F_{ST}$ values between each pair of wards.

In addition to measuring the between and within group cultural variance, the in-person survey will also elicit information regarding individual’s perceived costs/benefits of CoFMAs, as well as group costs/benefits of CoFMAs. We can then use the cultural $F_{ST}$ values, paired with perceived costs and benefits to assess whether or not group cooperation will be favored by selection using the Rogers model of group selection by selective migration, from hereon referred to as the Rogers Inequality.

The Rogers Inequality model was initially designed to measure genetic evolution of groups via selective migration, but is equally applicable to, and has been adapted for cultural evolution [@bowles2006group]. This model presents an inequality under which individually costly, yet group beneficial traits are favored by selection. The emergence of group cooperative norms will be favored by selection when the ratio of between-group variance ($F_{ST}$) to within-group variance ($1-F_{ST}$) exceeds the ratio of individual costs of a behavior ($C_{w}$) to group level population benefit ($B_{p}$). Thus, if the following inequality is true, then individually costly, group beneficial behavior is favored by selection [@kline2018designing].

$$\frac{c_{w}}{b_{p}} < \frac{F_{ST}}{(1-F_{ST})}$$

While the primary output of this model is a true/false estimate of whether the emergence of a costly cooperative norm is favored by selection, we can also use the ratio presented in the inequality as a measure of selection strength. For example, we will expect Community Forest Management Agreements to be more strongly selected for when the right-hand side of the inequality is far greater than the left-hand side compared to when the sides are close to equal.

## Preliminary Work
This project benefits from preliminary work done by our collaborators at the Max Planck Institute for Evolutionary Anthropology (MPI). MPI researchers conducted a preliminary qualitative survey and collation of oral histories with the assistance of Department of Forests and Non-Renewable Natural Resources staff to confirm anecdotally that competition between wards has led to CoFMA adoption at the ward level. These researchers have also visited wards across the island and recorded which wards have adopted CoFMAs and which have not. These adoption characteristics will be compared against the adoption characteristics recorded in my work as a form of data validation.

## Expected Outputs
The primary outputs of data collection and analysis in this project will produce binary predictions of protected area adoption for each study ward, as well as the calculated strength of selection for protected area adoption for each study ward. I will compare these calculated adoption likelihoods and selection strengths against actual adoption rates of CoFMAs. Figure \@ref(fig:selPress) shows an example of what the results of this study might look like. The findings of this study will be submitted for publication in a peer reviewed, academic journal.

```{r selPress, echo=FALSE, fig.cap="Simulated data showing protected area adoption vs calculated selection pressure for group adoption of protected areas.", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("Figures/Ch3/SelPressureAdopt.png")
```
\newpage

# Chapter 3: Social influences on transmission fidelity in household level diffusion of beneficial technology.

@@

This dissertation will achieve this goal in three progressive steps. The first will formalize theory around cultural group selection in conservation contexts. This step is critical as conservation is a fundamentally different system from where these theories were originally developed. The second step will empirically validate these theoretical insights by comparing calculated cultural group selection pressures to real-world adoption of Community Forest Management Agreements on Pemba Island, Tanzania. Lastly, this project will make use of a conservation initiative currently spreading through Pemban communities to examine how behavioral efficiency may be altered by social contexts (e.g. communication frequency) as a behavior spreads throughout a population. All together, this research will allow us to better understand and possibly predict the emergence and effective spread of conservation norms within a population.

>>>>

--- 
title: Cultural Evolution of Conservation Behaviors Within and Between Communities
author: "Matt Clark"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: yes
  pdf_document:
    toc: yes
    extra_dependencies: "subfig"
  bookdown::html_document2:
    toc: yes
fontsize: 12pt
link-citations: yes
csl: AmJBot.csl
subtitle: Dissertation Proposal for Committee Review
bibliography: [References/Dissertationproposal.bib,References/NonZoteroRefs.bib]
---
```{r packages, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(rmarkdown)
library(bookdown)
library(formattable)
library(kableExtra)
library(dplyr)
library(magrittr)
library(prettydoc)
library(htmltools)
library(knitcitations)
library(devtools)
library(tidyverse)
library(scales)
library(extrafont)
```

@@

# Chapter 2: Effects of Inter-Group Resource Competition and Selection on Community Adoption of Costly Conservation Policy.

## Background
Human social structures are characterized by nested groups, ranging from a nuclear family unit to a national or even multi-national identity. Group behavioral norms can have profound impacts on the wellbeing of group members and their environment, far beyond that of a single individual. Cultural group selection theorizes, and to a small extent has empirically demonstrated, the mechanisms which establish and perpetuate group-level norms [@henrich_cultural_2004; @bell_culture_2009]. Most relevant to community conservation is the development of norms which are beneficial for the group, but cause an individual to incur some personal cost; we refer to these as costly cooperative behaviors [@molleman_cultural_2013]. Central to theories around the adoption of costly cooperative behaviors is the insight that selection acts on both individuals and groups. Individuals may engage in personally costly behavior if it is beneficial for their group, which they also depend on for their wellbeing [@handley_human_2020]. The degree to which an individual depends on each of their nested groups is directly proportional to the personal costs they will incur for that group [@okasha_multilevel_2004; @price1972extension]. To the extent that conservation is a collaborative dilemma, these insights from cultural group selection may be paramount to implementing effective community based conservation [@waring2015multilevel; @kline2018designing].

@@

REDD+ in Zanzibar was designed to give performance based payments to communities who successfully reduce deforestation [@burgess2010getting]. In theory, these immediate, promised payments should alter the ratio of individual costs to group benefits so that costly cooperation is favored under cultural group selection. While this project was implemented in only a select few wards (Figure \@ref(fig:moon)), this setup should also promote the spread of community conservation norms to adjacent wards through success-biased imitation [@boyd_transmission_2010].

The REDD+ program on Pemba helped establish CoFMAs in 18 wards on the island, some of which have been successful in slowing and even stopping deforestation (Figure \@ref(fig:ForestCoverMap)). The payments for the conservation efforts however were never given to communities. While this is ethically troublesome, it is interesting in light of the fact that ten additional wards on the island have subsequently elected to establish CoFMAs. This leaves open the question of why community conservation areas are spreading on the island while they are not expected to do so under cultural group selection. One explanation posited by Andrews and Borgerhoff Mulder [-@andrews_cultural_2018] actually hinges on a major critique of REDD+; leakage [@andam2008measuring].

Leakage [as defined by @andam2008measuring] is a form of land use spillover where an intervention, such as the CoFMAs established on Pemba by REDD+, has a negative, counteracting effect on the targeted outcome variable in another place. On Pemba, we have anecdotal testimonies that individuals in wards with CoFMAs have simply shifted their harvest pressure onto adjacent wards. While dubious for the efficacy of REDD+ writ large, on Pemba, leakage may actually be driving the spread of community based conservation observed in the ten new communities which elected to create CoFMAs post-REDD+ [@andrewsdoes]. Just as the performance based payments from REDD+ were designed to alter the individual/community cost/benefit matrix so that selection favors costly collaboration via immediate community payments, we hypothesize that leakage from wards with CoFMAs may also act to favor costly cooperative behavior as individuals must act to defend their access to resources they would not have otherwise.

@@

### Predictions
* P1: Reduced deforestation rates inside CoFMA conservation areas will be correlated with increased deforestation in adjacent wards.

* P2: In wards that have subsequently applied for CoFMAS, individuals who voted in favor of implementing the CoFMA will report leakage as their primary motivation.

## Research Design
This project has two main components, both of which benefit from preliminary work done by our collaborators already working in Pemba, and both of which will require additional on-the-ground data collection. The first component is to assess if individuals adhering to CoFMA regulations is associated with accelerated forest loss in nearby wards (leakage). The second component is to assess people’s motivations for electing to create CoFMAs in their own wards and see if those have changed over time.

*Assessing leakage*

Using a combination of Landsat 5, Landsat 7, and Landsat 8 images our collaborators have created composite images of average forest cover across all of Pemba for the years 2001, 2010, and 2018 [@cohen2004landsat]. Data produced from these images classifies forest vs non-forest with approximately 95% accuracy across all three time periods [@andrewsdoes]. As shown in (Figure \@ref(fig:ForestCoverMap)), we have shapefiles for the individual wards on Pemba [accessed from GADM -@areas2020gadm] and can compare rates of forest loss for each ward across years. While this data is useful, it does not allow us to differentiate between forest loss inside or outside the demarcated conservation and use areas for each CoFMA. These data may then be skewed by misinterpreting tree planting for woodlots as reforestation, or forest loss inside designated use areas as deforestation.

To better assess if individuals are adhering to CoFMA regulations, we need to sort out the originally demarcated use and conservation areas in each CoFMA. To our knowledge, these data do not currently exist or are not available. As a first step in this research, we will attempt to access this information through the Department of Forestry and Non-Renewable Natural Resources libraries on Pemba. Where/if this data is not available, we will drive or walk the boundary of each area to create a shapefile to use for analysis.

We will compare the pre-CoFMA and post-CoFMA deforestation rates inside the conservation area for each ward to get a better measure of the degree to which individuals actually adhered to CoFMA regulations. To then estimate leakage, we will compare these values to the changes in deforestation rates between the two time periods for all ten of the wards that subsequently elected to create CoFMAs. If leakage did occur, we expect to see an association between the decline of deforestation inside CoFMA conservation areas and an increase in deforestation in nearby wards.

*Assessing motivations*

The ten wards which have subsequently elected to create CoFMAs did so in 2018 and 2019. We know from our collaborators that on the order of magnitude of approximately 500 individuals in each ward voted on the proposition. Our intention is to randomly sample individuals in each ward to gain a representative sample of each of these voting populations. By surveying these individuals, we hope to gain insights on two main questions. What was the reason they voted in favor of adopting a CoFMA if they did? And have their motivations changed?

To understand why individuals voted to adopt CoFMAs, we will ask about their perceptions of outside individuals stealing resources from their ward during that time period. We will ask a variety of questions about the individuals they believe were responsible and if they were aware of CoFMAs in adjacent wards. We will also ask a series of questions in regard to the long-term benefits of forest conservation in the absence of threats from outside groups. These questions will be targeted to let us tease apart the alternative explanations for why individuals voted to adopt CoFMAs.

To further triangulate on the reason(s) why individuals adopt community conservation areas, we will rely on the local, anecdotal insight that in the absence of REDD+ payments, some of the original CoFMAs are no longer being adhered to by individuals harvesting from the forest. If this is the case *and* leakage is causing the cascade of community conservation areas on Pemba, we expect that leakage pressure will have decreased in some CoFMA adjacent wards and individuals will no longer support subsequent creation of CoFMAs. In the survey, we will also ask about individual’s current willingness to adopt CoFMAs and their current perceptions of tree poaching and conservation benefits.

*Drawing conclusions*

The situation on Pemba in regard to REDD+ and community conservation areas is a bit of a natural experiment. There are 18 original wards with CoFMAs, each of those with a different proportion conservation-designated forest. Subsequently, each ward has shown different rates in forest cover loss and external wards have elected to create CoFMAs themselves. We hope to tie these disparate sources of information together to create a full picture of the social-ecological system.

We plan to analyze these results in a regression context where we will estimate the probability that an individual supports costly cooperation as a function of leakage. We plan to control for traits unique to individuals such as wealth and relative forest dependence. A final question we intend to answer is whether the hypothesized cascade of community conservation areas as a result of leakage is a permanent feature of the social-ecological landscape or just a temporary blip. The degree of drop off in support for CoFMAs between 2018/2019 and this research will serve as a good indicator for how the system may develop in the future.

## Preliminary Work
This project benefits from preliminary work done by our collaborators at the Max Planck Institute for Evolutionary Anthropology (MPI). MPI researchers conducted a preliminary qualitative survey and collation of oral histories with the assistance of Department of Forests and Non-Renewable Natural Resources staff to confirm anecdotally that leakage is occurring and has led to CoFMA adoption at the ward level. These researchers have also visited wards across the island and recorded which wards have adopted CoFMAs and which have not. This is in combination with the work mentioned above that our MPI collaborators have done on measuring deforestation rates on the island.

## Expected Outputs
This research is designed to contribute to our understanding of how and why community conservation areas spread and persist, and the adoption of costly cooperative norms broadly. Community buy-in and social-ecological outcomes of conservation areas vary significantly in practice [@naughton-treves_role_2005]. Agrawal and Redford [-@agrawal2006poverty] argue that this is because conservation areas are often created as “a shot in the dark,” with little attention paid to the mechanisms which cause conservation to be widely adopted and subsequently successful. This research will help close the gap in our understanding of conservation adoption and help us better allocate limited resources for conservation. The findings of this study will be submitted for publication in a peer reviewed, academic journal. 
Lastly, the findings of this research will help to empirically validate the insights from the agent-based model presented in the previous chapter. As we build validity in this model, we can use it to ask different and more in-depth questions about the nature of conservation adoption. The compatibility between this empirical study system and the agent-based model may allow us to generate and answer novel questions synergistically in a way neither project could do alone.

# Chapter 3: Social Influences on Transmission Fidelity in Household Level Diffusion of Beneficial Technology.

@@

This dissertation will achieve this goal in three progressive steps. The first will formalize theory around cultural group selection in conservation contexts. This step is critical as conservation is a fundamentally different system from where these theories were originally developed. The second step will empirically validate these theoretical insights by determining why community adoption of conservation areas is spreading on Pemba Island, Tanzania. Lastly, this project will make use of a conservation initiative currently spreading through Pemban communities to examine how behavioral efficiency may be altered by social contexts (e.g. communication frequency) as a behavior spreads throughout a population. All together, this research will allow us to better understand and possibly predict the emergence and effective spread of conservation norms within a population.
